FraCaS: Temporal Analysis

### 2020-06-10 paper2020/paper.tex
example: commentary

\section{Worked out example}
To give a sense of the additional details necessary to deal with the
precision demanded by a proof-assistant such as Coq we show how
problem (279) is worked out fully.

Input from \citet{Ljunglof:2012}:

{\small
\begin{verbatim}
s_279_1_p=
sentence 
(useCl past pPos
 (predVP
   (usePN (lexemePN "smith_PN"))
   (advVP
     (complSlash
      (slashV2a (lexemeV2 "write_V2")))
     (detCN (detQuant indefArt numSg)
       (useN (lexemeN "novel_N"))))
       (lexemeAdv "in_1991_Adv")))
s_279_3_h=
sentence
(useCl past pPos
 (predVP (usePN
   (lexemePN "smith_PN"))
   (advVP
     (complSlash
      (slashV2a (lexemeV2 "write_V2"))
      (usePron it_Pron))
     (lexemeAdv "in_1992_Adv"))))
\end{verbatim}
Interpretation by our dynamic and temporal semantics:

\begin{verbatim}
FORALL (fun a=>novel_N a)
(fun a=>(exists (b: Time),
((exists (c: Time),
(IS_INTERVAL Date_19910101 b /\
 IS_INTERVAL c Date_19911231 /\
 IS_INTERVAL b c /\
 appTime b c (write_V2 a)
   (PN2object smith_PN))))) ->
Not (exists (f: Time),
((exists (g: Time),
 (IS_INTERVAL Date_19920101 f /\
  IS_INTERVAL g Date_19921231 /\
  IS_INTERVAL f g /\
  appTime f g (write_V2 a) 
    (PN2object smith_PN)))))).
\end{verbatim}

Unfolding definitions at the Coq level:
\begin{verbatim}
forall x : object,
novel_N x ->
(exists b c : Z,
   Date_19910101 <= b /\
   c <= Date_19911231 /\
   b <= c /\ write_V2 x SMITH b c) ->
(exists f g : Z,
   Date_19920101 <= f /\
   g <= Date_19921231 /\
   f <= g /\ write_V2 x SMITH f g) ->
False
\end{verbatim}
(Very close to \cref{eq:smith-novel-1})

Proof:
\begin{verbatim}
Theorem  problem279 : Problem279aFalse.
cbv.
intros novel isSmithsNovel P1 H.
destruct P1 as
   [t0 [t1 [ct1 [ct2 [ct3 P1]]]]].
destruct H as
   [u0 [u1 [cu1 [cu2 [cu3 H]]]]].
specialize writeUnique
  with (x := novel)(y := SMITH) as A.
unfold UniqueActivity in A.
specialize (A _ _ _ _ P1 H) as B.
lia.
Qed.
\end{verbatim}
}

Critical: UniqueActivity; lia does arithmetic work.

>>>>

\section{Worked out example}
To give a sense of the additional details necessary to deal with the
precision demanded by a proof-assistant such as Coq we show how
problem (279) is worked out in full details.

We start with input trees in GF format, given by
\citet{Ljunglof:2012}. They can be rendered as follows:

{\small
\begin{verbatim}
s_279_1_p=
sentence 
(useCl past pPos
 (predVP
   (usePN (lexemePN "smith_PN"))
   (advVP
     (complSlash
      (slashV2a (lexemeV2 "write_V2")))
     (detCN (detQuant indefArt numSg)
       (useN (lexemeN "novel_N"))))
       (lexemeAdv "in_1991_Adv")))
s_279_3_h=
sentence
(useCl past pPos
 (predVP (usePN
   (lexemePN "smith_PN"))
   (advVP
     (complSlash
      (slashV2a (lexemeV2 "write_V2"))
      (usePron it_Pron))
     (lexemeAdv "in_1992_Adv"))))
\end{verbatim}

  Of particular note is the use of the pronoun ``it'', and the fact
  that adverbials expressions such that ``in 1992'' are lexicalized.
  We also follow the GF convention to postfix lexical items with the
  name of their category.

Our dynamic and temporal semantics gives the following interpretation
for \verb!s_279_1_p! implies \verb!s_279_3_h!.
\begin{verbatim}
FORALL (fun a=>novel_N a)
(fun a=>(exists (b: Time),
((exists (c: Time),
(IS_INTERVAL Date_19910101 b /\
 IS_INTERVAL c Date_19911231 /\
 IS_INTERVAL b c /\
 appTime b c (write_V2 a)
   (PN2object smith_PN))))) ->
Not (exists (f: Time),
((exists (g: Time),
 (IS_INTERVAL Date_19920101 f /\
  IS_INTERVAL g Date_19921231 /\
  IS_INTERVAL f g /\
  appTime f g (write_V2 a) 
    (PN2object smith_PN)))))).
\end{verbatim}
In the above, one should remark the top-level quantification over the
novel (as explained in \cref{sec:our-semantics}), the quantification
over time intervals as individual timepoints, and the use of custom
operators for several constructions (FORALL, Not, IS_INTERVAL,
appTime). This use of custom operators is useful for several
generalisations (for example, we have quantifiers such as MOST in
addition to FORALL.)

Unfolding the definitions for these operators yield the following
proposition:
\begin{verbatim}
forall x : object,
novel_N x ->
(exists b c : Z,
   Date_19910101 <= b /\
   c <= Date_19911231 /\
   b <= c /\ write_V2 x SMITH b c) ->
(exists f g : Z,
   Date_19920101 <= f /\
   g <= Date_19921231 /\
   f <= g /\ write_V2 x SMITH f g) ->
False
\end{verbatim}
This is very close to our idealised representation of the problem
\cref{eq:smith-novel-1}. One difference is the use of abstract Coq
integers for timepoints. Using a discrete time allows us to use
predifined Coq tactics. The discrete nature of integers does not
interfere with the reasoning.

Finally, we can show a Coq proof for the above proposition:
\begin{verbatim}
Theorem  problem279 : Problem279aFalse.
cbv.
intros novel isSmithsNovel P1 H.
destruct P1 as
   [t0 [t1 [ct1 [ct2 [ct3 P1]]]]].
destruct H as
   [u0 [u1 [cu1 [cu2 [cu3 H]]]]].
specialize writeUnique
  with (x := novel)(y := SMITH) as A.
unfold UniqueActivity in A.
specialize (A _ _ _ _ P1 H) as B.
lia.
Qed.
\end{verbatim}
}

The intros and destruct tactics serve bookkeeping purposes. The
critical part is the use of the \texttt{writeUnique} axiom, which
witnesses the aspectual class of the predicate \texttt{write\_V2}.
The proof is completed by the use of the \texttt{lia} tactic, which is
embeds a decision procedure for linear arithmetic
problems. Fortunately, \texttt{lia} can take care of all the problems
which arise in the FraCaS testsuite.
